---
title: "Pima Indians Diabetes"
output: 
  pdf_document:
    latex_engine: xelatex
author: ""
date: "`r Sys.Date()`"
---

## 1. Data loading

```{r chunk1}
# If the mlbench package is not installed yet, please install it first

library(mlbench)

# load
data("PimaIndiansDiabetes")

pima_data <- PimaIndiansDiabetes
str(pima_data)
```

```{r chunk1.1}
head(pima_data)
```



> **Description**
> The dataset contains 768 rows and 9 columns, and the main attributes include:
>
> * **pregnant**: number of pregnancies
> * **glucose**: plasma glucose concentration
> * **pressure**: diastolic blood pressure (mm Hg)
> * **triceps**: triceps skinfold thickness (mm)
> * **insulin**: 2-hour serum insulin (mu U/ml)
> * **mass**: BMI (body mass index)
> * **pedigree**: diabetes family history function
> * **age**: age
> * **diabetes**: target variable, factor category ("pos" means diabetes, "neg" means no disease)


## 2. cleaning and data preparing

### 2.1 NA

```{r chunk2}
# View basic statistics for each variable
summary(pima_data)

# Check the number of missing values
sapply(pima_data, function(x) sum(is.na(x)))
```


### 2.2 Dealing with outliers

A simple check on the glucose variable

```{r handle-zeros}
# Check the number of 0 values in glucose
sum(pima_data$glucose == 0)

# Cannot be 0
pima_data$glucose <- ifelse(pima_data$glucose == 0, NA, pima_data$glucose)

# View the updated missing values
table(is.na(pima_data$glucose))
```
```{r handle-zeros.1}
# the mean of glucose
glucose_mean <- mean(pima_data$glucose, na.rm = TRUE)

# substitute
pima_data$glucose[is.na(pima_data$glucose)] <- glucose_mean
```

```{r handle-zeros-rest}
# List of variables to clean (where 0 is likely invalid)
vars_to_clean <- c("pressure", "triceps", "insulin", "mass")

# Loop over variables
for (var in vars_to_clean) {
  # Convert 0 to NA
  pima_data[[var]] <- ifelse(pima_data[[var]] == 0, NA, pima_data[[var]])
  
  # Compute mean without NA
  var_mean <- mean(pima_data[[var]], na.rm = TRUE)
  
  # Replace NA with mean
  pima_data[[var]][is.na(pima_data[[var]])] <- var_mean
}
```

### 2.3 preprocessing

```{r preprocessing}
# For the target variable diabetes, confirm that it is a factor (if it is not already a factor)
pima_data$diabetes <- as.factor(pima_data$diabetes)

```


## 3.  Feature Interaction Analysis

### 3.1  Distribution of Numerical Features

```{r univariate-plots, fig.width=7, fig.height=5}
library(ggplot2)

# Plot histogram of age
ggplot(pima_data, aes(x = age)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Count")

# Plot histogram of glucose
ggplot(pima_data, aes(x = glucose)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Glucose", x = "Plasma Glucose Concentration", y = "Count")

```

### 3.2 Categorical Feature Impact on Target

```{r categorical-analysis}
# Frequency of different pregnancy counts
table(pima_data$pregnant)

# Analyze the effect of number of pregnancies on diabetes outcome
ggplot(pima_data, aes(x = as.factor(pregnant), fill = diabetes)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Pregnancy Count", 
       x = "Number of Pregnancies", 
       y = "Proportion", 
       fill = "Diabetes Status")

```
The sample size is obviously decreasing as the number of pregnancies increases.
It seems that the more pregnancies you have, the higher your risk of diabetes, but the data is not supportive at this time.

```{r p-analysis}
model <- glm(diabetes ~ pregnant, data = pima_data, family = binomial)
summary(model)
```
Although the sample size is small at very high pregnancies (e.g., ≥14), and the proportion graph may exaggerate the trend, overall, the logistic regression shows that the number of pregnancies is significantly positively associated with the incidence of diabetes. For each additional pregnancy, the odds of diabetes increase by approximately 14.7%.



### 3.3 Cross-Feature Interaction: Scatter Plots
```{r test0}
sum(is.na(pima_data$glucose))     
sum(is.na(pima_data$age))         
```

```{r scatter-analysis}
# Scatter plot: age vs. glucose colored by diabetes status
ggplot(pima_data, aes(x = age, y = glucose, color = diabetes)) +
  geom_point() +
  labs(title = "Age vs. Glucose Colored by Diabetes Status", 
       x = "Age", 
       y = "Plasma Glucose Concentration", 
       color = "Diabetes Status")

```
Among people with high blood sugar, the proportion of diabetes is higher, and the impact of age is not as significant as blood sugar




```{r boxplot1}
library(ggplot2)

# Boxplots for each variable
vars_to_plot <- c("pressure", "triceps", "insulin", "mass", "pedigree")

for (var in vars_to_plot) {
  p <- ggplot(pima_data, aes_string(x = "factor(diabetes)", y = var)) +
    geom_boxplot(fill = "skyblue") +
    labs(title = paste("Boxplot of", var, "by Diabetes Status"),
         x = "Diabetes (0 = No, 1 = Yes)",
         y = var)
  print(p)
}


```

The reason for the numerous outliers in **serum insulin** may be that insulin levels in the human body naturally fluctuate significantly.

I believe this feature may not contribute positively to the model's performance. Therefore, it's important to compare the results of models with and without this feature during training.

Triceps skinfold thickness is a commonly used anthropometric method for estimating body fat percentage. However, I also believe that this feature does not contribute positively to the model's performance.


```{r Density Plot}
for (var in vars_to_plot) {
  p <- ggplot(pima_data, aes_string(x = var, fill = "factor(diabetes)")) +
    geom_density(alpha = 0.4) +
    labs(title = paste("Density Plot of", var, "by Diabetes Status"),
         x = var, fill = "Diabetes") +
    theme_minimal()
  print(p)
}

```
Pressure: The overall blood pressure of diabetic patients is slightly higher. The difference is not significant.

Triceps: The two categories almost completely overlap, and the peak position in the middle is close.

Insulin: The data is very biased, with dense spikes and a large number of outliers.

Mass: The overall BMI of diabetic patients is higher.

Pedigree: People with high pedigree values are more likely to be ill.

```{r Facet Histogram}
for (var in vars_to_plot) {
  p <- ggplot(pima_data, aes_string(x = var)) +
    geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
    facet_wrap(~diabetes) +
    labs(title = paste("Histogram of", var, "by Diabetes Status"),
         x = var, y = "Count") +
    theme_minimal()
  print(p)
}


```


## 4. GLM Model Training

### 4.1 Train/Test Split

```{r train-test-split}

library(caret)


# Set seed for reproducibility
set.seed(123)

# Split dataset: 70% training, 30% testing
train_index <- createDataPartition(pima_data$diabetes, p = 0.7, list = FALSE)
train_data <- pima_data[train_index, ]
test_data <- pima_data[-train_index, ]
```


### 4.2 Train the Logistic Regression Model (GLM with Binomial Family)

```{r glm-training}
# Train logistic regression using all available features
glm_model <- glm(diabetes ~ ., data = train_data, family = "binomial")

# View model summary
summary(glm_model)
```
Based on the model output above, the non-significant variables include pressure, triceps, and insulin. These variables have p-values greater than 0.05.

```{r glm-training}
# Train logistic regression without the non-significant variables
diabetes_glm <- glm(diabetes ~ pregnant + glucose + mass + pedigree + age, 
                    data = train_data, 
                    family = "binomial")

# View summary of the simplified model
summary(diabetes_glm)
```

### 4.3 Prediction and Model Evaluation

```{r glm-prediction}
# Predict probabilities on the test set
pred_prob <- predict(glm_model, newdata = test_data, type = "response")

# Convert probabilities to class labels using a 0.5 threshold
pred_class <- ifelse(pred_prob > 0.5, "pos", "neg")
pred_class <- factor(pred_class, levels = levels(test_data$diabetes))

# Build confusion matrix
conf_matrix <- table(Predicted = pred_class, Actual = test_data$diabetes)
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
cat("Accuracy =", accuracy, "\n")
```



###  4.4 Evaluation Metrics: Precision, Recall, F1 Score, ROC AUC

```{r glm-metrics}
# Load required libraries
library(caret)
library(pROC)

# Precision
precision <- posPredValue(pred_class, test_data$diabetes, positive = "pos")

# Recall (also called Sensitivity)
recall <- sensitivity(pred_class, test_data$diabetes, positive = "pos")

# F1 Score
f1_score <- 2 * precision * recall / (precision + recall)

# Print metrics
cat("Precision:", round(precision, 3), "\n")
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1_score, 3), "\n")
```


### 4.5 ROC Curve and AUC (Area Under Curve)

```{r glm-roc-auc, message=FALSE}
# Convert true labels to numeric: pos = 1, neg = 0
actual_numeric <- ifelse(test_data$diabetes == "pos", 1, 0)

# Compute ROC
roc_obj <- roc(actual_numeric, pred_prob)

# Plot ROC Curve
plot(roc_obj, main = "ROC Curve for GLM", col = "blue")

# Calculate AUC
auc_value <- auc(roc_obj)
cat("AUC (Area Under the Curve):", round(auc_value, 3), "\n")
```

## Advanced Model Development and Training

### Data Preparation for Advanced Modeling

```{r advanced_setup}
# Load required libraries
library(caret)
library(randomForest)
library(pROC)
library(ggplot2)
library(dplyr)
library(VIM)
library(mlbench)

# Handle missing values (biologically implausible zeros) - if not already done
pima_clean <- pima_data
pima_clean$glucose[pima_clean$glucose == 0] <- NA
pima_clean$pressure[pima_clean$pressure == 0] <- NA
pima_clean$mass[pima_clean$mass == 0] <- NA

# Perform KNN imputation
pima_imputed <- kNN(pima_clean, k = 5)
pima_imputed <- pima_imputed[, 1:9]

# Convert diabetes to factor
pima_imputed$diabetes <- as.factor(pima_imputed$diabetes)

# Create interaction terms
pima_imputed$glucose_mass <- pima_imputed$glucose * pima_imputed$mass
pima_imputed$glucose_age <- pima_imputed$glucose * pima_imputed$age
pima_imputed$mass_age <- pima_imputed$mass * pima_imputed$age

# Create categorical variables
pima_imputed$age_group <- cut(pima_imputed$age, 
                              breaks = c(0, 30, 45, 60, 100), 
                              labels = c("Young", "Middle", "Senior", "Elderly"))

pima_imputed$glucose_group <- cut(pima_imputed$glucose, 
                                  breaks = c(0, 100, 126, 200, 1000), 
                                  labels = c("Normal", "Prediabetes", "Diabetes", "High"))

pima_imputed$mass_group <- cut(pima_imputed$mass, 
                               breaks = c(0, 18.5, 25, 30, 100), 
                               labels = c("Underweight", "Normal", "Overweight", "Obese"))

# Convert categorical variables to numeric for modeling
pima_imputed$age_group_num <- as.numeric(pima_imputed$age_group)
pima_imputed$glucose_group_num <- as.numeric(pima_imputed$glucose_group)
pima_imputed$mass_group_num <- as.numeric(pima_imputed$mass_group)

# Remove original categorical columns to keep dataset clean
pima_imputed <- pima_imputed[, !colnames(pima_imputed) %in% 
                             c("age_group", "glucose_group", "mass_group")]

cat("Total columns after feature engineering:", ncol(pima_imputed), "\n")
```

### Advanced Data Splitting

```{r advanced_splitting}
# Set seed for reproducibility
set.seed(123)

# Split data: 70% training, 15% validation, 15% test
train_index <- createDataPartition(pima_imputed$diabetes, p = 0.7, list = FALSE)
train_data <- pima_imputed[train_index, ]
temp_data <- pima_imputed[-train_index, ]

val_index <- createDataPartition(temp_data$diabetes, p = 0.5, list = FALSE)
val_data <- temp_data[val_index, ]
test_data <- temp_data[-val_index, ]

# Check split sizes and class distribution
cat("Training set size:", nrow(train_data), "\n")
cat("Validation set size:", nrow(val_data), "\n")
cat("Test set size:", nrow(test_data), "\n")

cat("\nClass distribution in training set:\n")
print(table(train_data$diabetes))
cat("\nClass distribution in validation set:\n")
print(table(val_data$diabetes))
cat("\nClass distribution in test set:\n")
print(table(test_data$diabetes))
```

### Feature Scaling

```{r advanced_scaling}
# Identify numerical features (excluding target)
numerical_features <- c("pregnant", "glucose", "pressure", "triceps", 
                       "insulin", "mass", "pedigree", "age",
                       "glucose_mass", "glucose_age", "mass_age",
                       "age_group_num", "glucose_group_num", "mass_group_num")

# Calculate scaling parameters from training data
preprocess_params <- preProcess(train_data[, numerical_features], 
                               method = c("center", "scale"))

# Apply scaling to all datasets
train_scaled <- train_data
val_scaled <- val_data
test_scaled <- test_data

train_scaled[, numerical_features] <- predict(preprocess_params, train_data[, numerical_features])
val_scaled[, numerical_features] <- predict(preprocess_params, val_data[, numerical_features])
test_scaled[, numerical_features] <- predict(preprocess_params, test_data[, numerical_features])
```

### Advanced Model Training

```{r advanced_training}
# Set up cross-validation control
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE
)

# 1. Logistic Regression
set.seed(123)
lr_model <- train(
  diabetes ~ .,
  data = train_scaled,
  method = "glm",
  trControl = ctrl,
  metric = "ROC"
)

# 2. Random Forest
set.seed(123)
rf_model <- train(
  diabetes ~ .,
  data = train_scaled,
  method = "rf",
  trControl = ctrl,
  metric = "ROC",
  tuneLength = 3
)

# 3. XGBoost (Simplified version)
set.seed(123)
xgb_model <- train(
  diabetes ~ .,
  data = train_scaled,
  method = "xgbTree",
  trControl = ctrl,
  metric = "ROC",
  tuneLength = 3,  # Reduced tuning
  verbose = FALSE  # Reduce output
)
```

### Model Comparison

```{r advanced_comparison}
# Compare model performances
models <- list(
  "Logistic Regression" = lr_model,
  "Random Forest" = rf_model,
  "XGBoost" = xgb_model
)

# Extract results
results <- resamples(models)
summary(results)

# Plot comparison
bwplot(results, metric = "ROC")
dotplot(results, metric = "ROC")
```

### Hyperparameter Tuning

```{r advanced_tuning}
# Random Forest with detailed tuning
rf_grid <- expand.grid(
  mtry = c(2, 3, 4, 5, 6, 7, 8)
)

set.seed(123)
rf_tuned <- train(
  diabetes ~ .,
  data = train_scaled,
  method = "rf",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = rf_grid
)
```

### Advanced Model Evaluation

```{r advanced_evaluation}
# Function to evaluate model performance
evaluate_model <- function(model, test_data, model_name) {
  # Make predictions
  predictions <- predict(model, test_data, type = "prob")
  predicted_class <- predict(model, test_data)
  
  # Get the positive class name
  positive_class <- levels(test_data$diabetes)[2]  # Second level is usually positive
  
  # Calculate metrics
  cm <- confusionMatrix(predicted_class, test_data$diabetes, positive = positive_class)
  
  # Use the correct probability column name
  if (positive_class %in% colnames(predictions)) {
    roc_obj <- roc(test_data$diabetes, predictions[[positive_class]])
  } else {
    # Use the second column (usually the positive class)
    roc_obj <- roc(test_data$diabetes, predictions[, 2])
  }
  auc_score <- auc(roc_obj)
  
  # Return results
  results <- list(
    model_name = model_name,
    accuracy = cm$overall["Accuracy"],
    sensitivity = cm$byClass["Sensitivity"],
    specificity = cm$byClass["Specificity"],
    precision = cm$byClass["Precision"],
    f1_score = cm$byClass["F1"],
    auc = auc_score,
    confusion_matrix = cm,
    roc_obj = roc_obj
  )
  
  return(results)
}

# Evaluate working models on validation set
validation_results <- list()

validation_results$lr <- evaluate_model(lr_model, val_scaled, "Logistic Regression")
validation_results$rf <- evaluate_model(rf_tuned, val_scaled, "Random Forest")
validation_results$xgb <- evaluate_model(xgb_model, val_scaled, "XGBoost")

# Create comparison table
comparison_table <- data.frame(
  Model = sapply(validation_results, function(x) x$model_name),
  Accuracy = sapply(validation_results, function(x) x$accuracy),
  Sensitivity = sapply(validation_results, function(x) x$sensitivity),
  Specificity = sapply(validation_results, function(x) x$specificity),
  Precision = sapply(validation_results, function(x) x$precision),
  F1_Score = sapply(validation_results, function(x) x$f1_score),
  AUC = sapply(validation_results, function(x) x$auc)
)

print(comparison_table)
```

### ROC Curves

```{r advanced_roc}
# Plot ROC curves (only working models)
plot(validation_results$lr$roc_obj, col = "blue", main = "ROC Curves Comparison")
plot(validation_results$rf$roc_obj, col = "red", add = TRUE)
plot(validation_results$xgb$roc_obj, col = "purple", add = TRUE)
legend("bottomright", 
       legend = c("Logistic Regression", "Random Forest", "XGBoost"),
       col = c("blue", "red", "purple"), 
       lty = 1)
```

### Feature Importance

```{r advanced_importance}
# Feature importance for Random Forest
rf_importance <- varImp(rf_tuned)
plot(rf_importance, main = "Feature Importance - Random Forest")

# Feature importance for XGBoost
xgb_importance <- varImp(xgb_model)
plot(xgb_importance, main = "Feature Importance - XGBoost")

# Print importance scores
print(rf_importance)
print(xgb_importance)
```

### Final Model Selection

```{r advanced_final}
# Select best model based on validation performance
best_model_name <- comparison_table$Model[which.max(comparison_table$AUC)]
cat("Best model based on AUC:", best_model_name, "\n")

# Get the best model
if (best_model_name == "Logistic Regression") {
  best_model <- lr_model
} else if (best_model_name == "Random Forest") {
  best_model <- rf_tuned
} else if (best_model_name == "XGBoost") {
  best_model <- xgb_model
}

# Evaluate best model on test set
final_results <- evaluate_model(best_model, test_scaled, paste("Best Model:", best_model_name))

cat("\nFinal Test Set Results:\n")
cat("Accuracy:", round(final_results$accuracy, 4), "\n")
cat("Sensitivity:", round(final_results$sensitivity, 4), "\n")
cat("Specificity:", round(final_results$specificity, 4), "\n")
cat("Precision:", round(final_results$precision, 4), "\n")
cat("F1 Score:", round(final_results$f1_score, 4), "\n")
cat("AUC:", round(final_results$auc, 4), "\n")

# Final confusion matrix
print(final_results$confusion_matrix)
```

### Model Interpretation

```{r advanced_interpretation}
# For logistic regression, show coefficients
if (best_model_name == "Logistic Regression") {
  coef_summary <- summary(best_model)$coefficients
  print("Logistic Regression Coefficients:")
  print(coef_summary)
  
  # Calculate odds ratios
  odds_ratios <- exp(coef_summary[, "Estimate"])
  
  # Calculate confidence intervals manually (since confint doesn't work with caret objects)
  std_errors <- coef_summary[, "Std. Error"]
  z_value <- 1.96  # 95% confidence interval
  
  lower_ci <- exp(coef_summary[, "Estimate"] - z_value * std_errors)
  upper_ci <- exp(coef_summary[, "Estimate"] + z_value * std_errors)
  
  odds_table <- data.frame(
    Feature = rownames(coef_summary),
    Coefficient = coef_summary[, "Estimate"],
    Odds_Ratio = odds_ratios,
    Lower_CI = lower_ci,
    Upper_CI = upper_ci,
    P_value = coef_summary[, "Pr(>|z|)"]
  )
  
  print("Odds Ratios and Confidence Intervals:")
  print(odds_table)
}
```

### Summary and Conclusions

```{r advanced_summary}
# Summary statistics
cat("Pima Indians Diabetes Dataset Summary:\n")
cat("Total samples:", nrow(pima_imputed), "\n")
cat("Diabetes prevalence:", round(mean(pima_imputed$diabetes == "tested_positive"), 3), "\n")
cat("Class imbalance ratio:", round(sum(pima_imputed$diabetes == "tested_negative") / sum(pima_imputed$diabetes == "tested_positive"), 2), ":1\n")

cat("\nKey Findings:\n")

# Handle feature importance for different model types
if (best_model_name == "Logistic Regression") {
  # For logistic regression, use coefficients
  coef_importance <- abs(summary(best_model)$coefficients[, "Estimate"])
  top_features <- names(sort(coef_importance, decreasing = TRUE)[1:5])
  cat("1. Most important features:", paste(top_features, collapse = ", "), "\n")
} else {
  # For Random Forest, use variable importance
  tryCatch({
    rf_importance <- varImp(best_model)$importance
    # Extract the importance values as a vector
    if (is.data.frame(rf_importance)) {
      importance_values <- rf_importance[, 1]
    } else {
      importance_values <- as.numeric(rf_importance)
    }
    # Get feature names
    if (is.data.frame(rf_importance)) {
      feature_names <- rownames(rf_importance)
    } else {
      feature_names <- names(rf_importance)
    }
    # Sort and get top 5
    sorted_indices <- order(importance_values, decreasing = TRUE)
    top_features <- feature_names[sorted_indices[1:5]]
    cat("1. Most important features:", paste(top_features, collapse = ", "), "\n")
  }, error = function(e) {
    cat("1. Most important features: Unable to extract (using Random Forest)\n")
  })
}

cat("2. Best performing model:", best_model_name, "\n")
cat("3. Final AUC on test set:", round(final_results$auc, 4), "\n")
cat("4. Model sensitivity:", round(final_results$sensitivity, 4), "\n")
cat("5. Model specificity:", round(final_results$specificity, 4), "\n")

cat("\nComparison with Literature:\n")
cat("The Pima Indians Diabetes dataset is a classic benchmark in diabetes prediction.\n")
cat("Our results can be compared with numerous studies that have used this dataset.\n")
cat("Typical AUC scores range from 0.70 to 0.85 depending on the methodology used.\n")
```

